Reviewer Comments

### 2024-10-15 Sections/07-genetics.qmd
Further progress on review

Genetic data can be applied to complement ecological data to integrate evolutionary processes and patterns into landscape ecology [@Holderegger2008LandscapeGenetics; @Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics]. 
Landscape genetics unites molecular population genetics, spatial statistics, and landscape ecology, and emerged from the goal to study the interaction between landscape features and microevolutionary processes, such as gene flow, genetic drift and selection [@Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics]. 
Landscape genetic approaches have been used traditionally to inform landscape ecology applications, e.g., through connectivity modeling and spatial conservation planning [@Manel2003LandscapeGenetics; @Holderegger2008LandscapeGenetics; @Manel2013TenYears; @Bolliger2014LandscapeGenetics].
Using holistic approaches that consider evolutionary processes and patterns in addition to ecological data can fortify results, e.g., improving predictive models of species range shifts in response to climate change [e.g., @Aguirre-Liguori2021EvolutionaryGenomics], improving the identification and delineation of landscape connections among populations [e.g., @Homola2019LandscapeGenetics; @deFlamingh2024IntegratingHabitat], allowing for the interpretation of spatial structuring in context of socio-cultural connections [e.g., @Henson2021ConvergentGeographic], or understanding local adaptation associated with specific environments [@Sork2013PuttingLandscape; @Rolland2015CombiningNiche; @Aguirre-Liguori2021EvolutionaryGenomics].

Advances in molecular methodologies and  synergistic developments of bioinformatic and computational approaches to analyze large-scale genomic data have recently enabled the integration of genome-wide data with spatial ecological data.<!-- MH: Ref missing -->.
This shift towards landscape genomics [@Cushman2018EditorialLeast; @Storfer2018NavigatingInterface] comes with novel utilities through which researchers can apply genomic data across spatio-temporal scales, including the integration of historic or ancient DNA [@Fenderson2020SpatiotemporalLandscape], or simulated present-day and future population genomic data [@House2018EvaluatingMethods; @Chevy2024PopulationGenetics]. 
Analytical frameworks such as the *FOLDS* model [gene flow, genetic offsets, genetic load, dispersal and SDMs\; @Aguirre-Liguori2021EvolutionaryGenomics], constrained coordination where covarying sets of genotypes are correlated with multivariate environments through redundancy analysis [RDA\; @Capblancq2021RedundancyAnalysis], and spatial simulations provide promising potential for future landscape genomic studies. 
Such studies may explore interactions between complex evolutionary processes, including demography, multidimensional (e.g., strength and direction) gene flow and migration across spatio-temporal scales, genetic differentiation across landscapes, genomic load and adaptive potential, and the interpretation of these evolutionary processes in context of landscape ecology [@Balkenhol2017LandscapeGenomics; @House2018EvaluatingMethods; @Fenderson2020SpatiotemporalLandscape; @Marcus2021FastFlexible; @Chevy2024PopulationGenetics].
<!--JN: software??-->

>>>>

Genetic data can be applied in complement with ecological data to integrate evolutionary processes and patterns into landscape ecology [@Holderegger2008LandscapeGenetics; @Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics].
Landscape genetics unites molecular population genetics, spatial statistics, and landscape ecology, and emerged from the goal to study the interaction between landscape features and microevolutionary processes, such as gene flow, genetic drift and selection [@Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics].
Landscape genetic approaches have traditionally been used to inform landscape ecology applications, e.g., through connectivity modelling and spatial conservation planning [@Manel2003LandscapeGenetics; @Holderegger2008LandscapeGenetics; @Manel2013TenYears; @Bolliger2014LandscapeGenetics].
Using holistic approaches that consider evolutionary processes and patterns in addition to ecological data can fortify results, e.g., by improving predictive models of species range shifts in response to climate change [e.g., @Aguirre-Liguori2021EvolutionaryGenomics], improving the identification and delineation of landscape connections among populations [e.g., @Homola2019LandscapeGenetics; @deFlamingh2024IntegratingHabitat], allowing for the interpretation of spatial structuring in context of socio-cultural connections [e.g., @Henson2021ConvergentGeographic], or understanding local adaptation associated with specific environments [@Sork2013PuttingLandscape; @Rolland2015CombiningNiche; @Aguirre-Liguori2021EvolutionaryGenomics].

Advances in molecular methodologies and synergistic developments of bioinformatic and computational approaches to analyze large-scale genomic data have recently enabled the integration of genome-wide data with spatial ecological data.
This shift towards landscape genomics [@Cushman2018EditorialLeast; @Storfer2018NavigatingInterface] comes with novel utilities through which researchers can apply genomic data across spatio-temporal scales, including the integration of historic or ancient DNA [@Fenderson2020SpatiotemporalLandscape], or simulated present-day and future population genomic data [@House2018EvaluatingMethods; @Chevy2024PopulationGenetics].
Analytical frameworks such as the *FOLDS* model [gene flow, genetic offsets, genetic load, dispersal and SDMs\; @Aguirre-Liguori2021EvolutionaryGenomics], constrained coordination where covarying sets of genotypes are correlated with multivariate environments through redundancy analysis [RDA\; @Capblancq2021RedundancyAnalysis], and spatial simulations provide promising potential for future landscape genomic studies.
Such studies may explore interactions between complex evolutionary processes, including demography, multidimensional (e.g., strength and direction) gene flow and migration across spatio-temporal scales, genetic differentiation across landscapes, genomic load and adaptive potential, and the interpretation of these evolutionary processes in context of landscape ecology [@Balkenhol2017LandscapeGenomics; @House2018EvaluatingMethods; @Fenderson2020SpatiotemporalLandscape; @Marcus2021FastFlexible; @Chevy2024PopulationGenetics].

<!--
Given the importance of connectivity, all previously discussed connectivity software are also relevant for landscape genetics.
Additionally, popular software for landscape genetics includes the R packages *graph4lg* [@Savary2021Graph4lgPackage], *landgenreport* [@Gruber2015LandgenreportNew], *adegenet* [@Jombart2008AdegenetPackage], or *ape* [@Paradis2019Ape50], among others.
Furthermore, a series of Python scripts have been developed for use in landscape genetics studies [@Etherington2011PythonBased], or GIS-based toolboxes written in Python exist that provide landscape genetics functionality [e.g., @Brown2017SDMtoolbox20].
-->

### 2024-10-15 Sections/08-spatial-planning.qmd
Further progress on review

<!-- R1: Spatial conservation planning -> spatial planning -->
# Spatial conservation planning {#sec-spatialplanning}

@@

Particularly noteworthy are two advances that incorporate landscape metrics in spatial planning.
<!-- R1: Reference 204 relies on constraint programming, not linear programming; Justeau-Allaire2021ConstrainedOptimization -->
One is the use of graph theory in linear programming to guide conservation and restoration efforts [@Justeau-Allaire2021ConstrainedOptimization; @Hamonic2023CumulativeEffects].
These spatial networks approaches have the potential to provide more cost-effective and precise solutions to design reserve networks [@Hamonic2023CumulativeEffects].
For example, @Justeau-Allaire2021ConstrainedOptimization used landscape ecology theory and metrics to spatially optimize restoration efforts, aiming to maximize broader landscape connectivity in New Caledonia.

A second main development is the increasing use of deep reinforcement learning, a machine learning approach based not on patterns but agents and pathways.
<!-- R1: Beware, deep reinforcement learning is not an exact approach, and thus cannot provide guarantees that the best possible solution has been found. Its strength mainly lies in the fact that, for large problems where exact approaches are currently inefficient, it can approximate high-quality solutions -->
Reinforcement learning can be used to identify cost-efficient solutions to area-based conservation planning [@Silvestro2022ImprovingBiodiversity].
Additionally, it can be applied to identify the best possible solutions to improve landscape connectivity indices [@Equihua2024ConnectivityConservation], such as the integral index of connectivity [@Pascual-Hortal2006ComparisonDevelopment].
Advantages of reinforcement learning include the ease of incorporating both linear and non-linear functions, as well as the scalability of spatio-temporal processes to larger spatial extents [@Equihua2024ConnectivityConservation].
However, this comes with the drawback of reduced interpretability.

Multiple open-source software solutions have been developed to support integration of landscape ecology theory and indicators into spatial planning.
Most software are based on linear or constrained programming approaches and require a solver to create any outputs.
<!-- R1: Be careful with such statements, it is extremely complex to objectively compare two solvers, and almost impossible to state the general superiority of one state-of-the-art solver against another one. Indeed, some solvers may be very efficient for some kinds of problems while very inefficient for others. Also, modelling and solver parametrization can make a lot of differences in solving efficiency and has often a lot of impact on performances. They are solver competitions, however, they are based on benchmarks, therefore classic and well-known problems that do not always reflect common uses of solvers. A solver fine-tuned to be good on such benchmarks will not necessarily perform well on any problem. There are many excellent open-source solvers (in MILP, CP, and SAT), and their openness offers a lot more possibilities for extension and fine-tuning to specific problems -->
Unfortunately there is still a gap between open-source and proprietary solvers in terms of computational efficiency [@Hanson2019OptimalityPrioritizing].

The *prioritizr* R package [@Hanson2024PrioritizrSystematic] supports a range of compactness and contiguity constraints, penalties, and generally allows flexibility to create planning problems in multiple configurations.
Similarly, the *coco* R package uses connectivity constraints in an integer linear planning framework to identify potential outcomes that minimize further fragmentation of the landscape [@vanMantgem2023CocoConservation].
The *restoptr* R package uses constrained programming to identify restoration areas that would most effectively contribute to improved landscape connectivity [@Justeau-Allaire2023RestoptrPackage].
Innovatively, it uses landscape ecology indices not as constraints, but as specific targets, e.g., where one could optimize the effective mesh size in a region.
The *CAPTAIN* Python software library implements reinforcement learning for area-based conservation efforts [@Silvestro2022ImprovingBiodiversity].
However, many spatial planning software still require substantial prior knowledge on spatial optimization and statistics, which can create a barrier to more widespread adoption.

>>>>

<!-- R1: Spatial conservation planning -> spatial planning -->
# Spatial planning {#sec-spatialplanning}

@@

Particularly noteworthy are two advances that incorporate landscape metrics in spatial planning.
<!-- R1: Reference 204 relies on constraint programming, not linear programming; Justeau-Allaire2021ConstrainedOptimization -->
One is the use of graph theory in linear [@Hamonic2023CumulativeEffects] or constraint programming [@Justeau-Allaire2021ConstrainedOptimization] to guide conservation and restoration efforts.
These spatial networks approaches have the potential to provide more cost-effective and precise solutions to design reserve networks [@Hamonic2023CumulativeEffects].
For example, @Justeau-Allaire2021ConstrainedOptimization used landscape ecology theory and metrics to spatially optimize restoration efforts, aiming to maximize broader landscape connectivity in New Caledonia.

A second main development is the increasing use of deep reinforcement learning, a machine learning approach based not on patterns but agents and pathways.
<!-- R1: Beware, deep reinforcement learning is not an exact approach, and thus cannot provide guarantees that the best possible solution has been found. Its strength mainly lies in the fact that, for large problems where exact approaches are currently inefficient, it can approximate high-quality solutions -->
Reinforcement learning can be used to identify cost-efficient solutions to area-based conservation planning [@Silvestro2022ImprovingBiodiversity].
Additionally, it can be applied to identify the best possible solutions to improve landscape connectivity indices [@Equihua2024ConnectivityConservation], such as the integral index of connectivity [@Pascual-Hortal2006ComparisonDevelopment].
Advantages of reinforcement learning include the ease of incorporating both linear and non-linear functions, as well as the scalability of spatio-temporal processes to larger spatial extents [@Equihua2024ConnectivityConservation].
However, this comes with the drawbacks of reduced interpretability and no guarantees that the best possible solution can been found because it is not an exact approach.

Multiple open-source software solutions have been developed to support integration of landscape ecology theory and indicators into spatial planning.
Most software are based on linear or constrained programming approaches and require a solver to create any outputs.
<!-- R1: Be careful with such statements, it is extremely complex to objectively compare two solvers, and almost impossible to state the general superiority of one state-of-the-art solver against another one. Indeed, some solvers may be very efficient for some kinds of problems while very inefficient for others. Also, modelling and solver parametrization can make a lot of differences in solving efficiency and has often a lot of impact on performances. They are solver competitions, however, they are based on benchmarks, therefore classic and well-known problems that do not always reflect common uses of solvers. A solver fine-tuned to be good on such benchmarks will not necessarily perform well on any problem. There are many excellent open-source solvers (in MILP, CP, and SAT), and their openness offers a lot more possibilities for extension and fine-tuning to specific problems -->
Unfortunately there are differences between open-source and proprietary solvers in terms of computational efficiency [@Hanson2019OptimalityPrioritizing], and the optimal solver will depend on the specific problem. 
However, due to software licenses and related cost, the optimal solver might not be always accessible.

<!--
The *prioritizr* R package [@Hanson2024PrioritizrSystematic] supports a range of compactness and contiguity constraints, penalties, and generally allows flexibility to create planning problems in multiple configurations.
Similarly, the *coco* R package uses connectivity constraints in an integer linear planning framework to identify potential outcomes that minimize further fragmentation of the landscape [@vanMantgem2023CocoConservation].
The *restoptr* R package uses constrained programming to identify restoration areas that would most effectively contribute to improved landscape connectivity [@Justeau-Allaire2023RestoptrPackage].
Innovatively, it uses landscape ecology indices not as constraints, but as specific targets, e.g., where one could optimize the effective mesh size in a region.
The *CAPTAIN* Python software library implements reinforcement learning for area-based conservation efforts [@Silvestro2022ImprovingBiodiversity].
However, many spatial planning software still require substantial prior knowledge on spatial optimization and statistics, which can create a barrier to more widespread adoption.
-->

### 2024-10-15 Sections/09-conclusion.qmd
Further progress on review

>>>>

<!-- MH: Check connectivity, simulations sections again for 'hidden' software -->

| Topic | Programming language | Software | Reference |
|:---:|:---:|:---:|:---:|
| Spatial patterns | R | sf | [@Pebesma2018SimpleFeatures] |
| Spatial patterns | R | terra | [@Hijmans2024TerraSpatial] |
| Spatial patterns | R | landscapemetrics | [@Hesselbarth2019LandscapemetricsOpensource] |
| Spatial patterns | R | multilandr | [@Huais2024MultilandrPackage] |
| Spatial patterns | R | bespatial | [@Nowosad2024BespatialBoltzmann] |
| Spatial patterns | R | rasterdiv | [@Rocchini2021RasterdivInformation] |
| Spatial patterns | R | LandComp | [@Konrad2023AdaptingMultiscale] |
| Spatial patterns | R | geodiv | [@Smith2021GeodivPackage] |
| Spatial patterns | R | glcm | [@Zvoleff2020GlcmCalculate] |
| Spatial patterns | R | GraySpatCon | [@Riitters2023MappingLandscape] |
| Spatial patterns | R | vectormetrics | [@Matuszek2024VectormetricsLandscape] |
| Spatial patterns | R | motif | [@Nowosad2021MotifOpensource] |
| Spatial patterns | Python | GeoPandas | [@Jordahl2020GeopandasPython] |
| Spatial patterns | Python | rasterio | [@Gillies2013RasterioGeospatial] |
| Spatial patterns | Python | PyLandStats | [@Bosch2019PyLandStatsOpensource] |
| Spatial patterns | Julia | GeoStats | [@Hoffimann2018GeoStatsjlHighperformance] |
| Spatial patterns | Julia | GeoInterface | [@JuliaGeoandcontributors2024GeoInterfaceJulia] |
| Spatial patterns | Stand-alone software | FRAGSTATS | [@McGarigal2023FRAGSTATSV4] |
| Spatial patterns | Stand-alone software | GuidosToolbox | [@Vogt2017GuidosToolboxUniversal] |
| Connectivity | R | ResistanceGA | [@Peterman2018ResistanceGAPackage] |
| Connectivity | R | lconnect | [@Mestre2023LconnectPackage] |
| Connectivity | R | grainscape | [@Chubaty2020ToolboxGrainscape] |
| Connectivity | Julia | Circuitscape | [@Anantharaman2019CircuitscapeJulia] |
| Connectivity | Julia | ConScape | [@vanMoorter2023AcceleratingAdvances] |
| Connectivity | Julia | Omniscape | [@Landau2021OmniscapejlSoftware] |
| Connectivity | Stand-alone software | Graphab | [@Foltete2012SoftwareTool] |
| Connectivity | Stand-alone software | Conefor | [@Saura2009ConeforSensinode] |
| Connectivity | Stand-alone software | lscorridors | [@Ribeiro2017LandScapeCorridors] |
| Sampling | Python | ZonalMetrics | [@Adamczyk2017ZonalMetricsPython] |
| Simulations | R | NLMR | [@Sciaini2018NLMRLandscapetools] |
| Simulations | Python | NLMpy | [@Etherington2015NLMpyPython] |
| Simulations | Julia | NeutralLandscapes | [@Poisot2023NeutralLandscapesJulia] |
| Simulations | Stand-alone software | GradientLand | [@Cambui2015GradientLandSoftware] |
| Simulations | Stand-alone software | HexSim | [@Schumaker2018HexSimModeling] |
| Genetics | R | graph4lg | [@Savary2021Graph4lgPackage] |
| Genetics | R | landgenreport | [@Gruber2015LandgenreportNew] |
| Genetics | R | adegenet | [@Jombart2008AdegenetPackage] |
| Genetics | R | ape | [@Paradis2019Ape50] |
| Genetics | Python | Python based scripts | [@Etherington2011PythonBased] |
| Genetics | Python | SDMtoolbox 2.0 | [@Brown2017SDMtoolbox20] |
| Planning | R | prioritizr | [@Hanson2024PrioritizrSystematic] |
| Planning | R | coco | [@vanMantgem2023CocoConservation] |
| Planning | R | restoptr | [@Justeau-Allaire2023RestoptrPackage] |
| Planning | Python | CAPTAIN | [@Silvestro2022ImprovingBiodiversity] |

: Software implementing methodological developments in landscace ecology.
